---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BAtrending

<!-- badges: start -->
<!-- badges: end -->

A package for conveniently conducting a Bland-Altman including assessment of trending ability.

## Installation

You can install BAtrending from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("JohannesNE/BAtrending")
```

or 

``` r
devtools::install_github("JohannesNE/BAtrending")
```

## Example

BAtrending package includes a small sample dataset, `CO`. It contains paired measurements of cardiac output (CO) with two methods: radionuclide ventriculography (rv) and impedance cardiography (ic). The dataset was published in [Bland JM, Altman DG. (1999) Measuring agreement in method comparison studies. Statistical Methods in Medical Research 8, 135-160](https://doi.org/10.1177/096228029900800204).

The `CO` dataset has 60 measurements in 12 subjects.

```{r example, cache = TRUE}
library(BAtrending)

compare_CO <- compare_methods(CO, 
                              ref_col = "rv", 
                              alt_col = "ic", 
                              id_col = "sub")
# Bootstrap confidence intervals
compare_CO <- add_confint(compare_CO)

plot_BA_combine(compare_CO, fix_aspect_ratio = TRUE)
```

The object (`compare_CO`) returned by `compare_methods()` contains a number of parameters from both the standard Bland-Altman analysis and for assessing trending ability. The `print()` method gives an overview: 

```{r, R.options = list(width = 300)}
compare_CO
```

## Citation

```{r, results = 'asis'}
citation("BAtrending")
```
