---
title: "Introduction to BAtrending"
output: 
  rmarkdown::html_vignette: default
  pdf_document:
   toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to BAtrending}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>"
)
```

# Setup

First, install the package using `pak::pak("JohannesNE/BAtrending")` or `remotes::install_github("JohannesNE/BAtrending")`. Then, load the package.

```{r setup}
library(BAtrending)
```

# An Example

BAtrending package includes a small sample dataset, `CO`. It contains paired measurements of cardiac output (CO) with two methods: radionuclide ventriculography (rv) and impedance cardiography (ic). The dataset was published in [Bland JM, Altman DG. (1999) Measuring agreement in method comparison studies. Statistical Methods in Medical Research 8, 135-160](https://doi.org/10.1177/096228029900800204).

The `CO` dataset has 60 measurements in 12 subjects.

```{r}
head(CO)
```

`compare_methods()` is the primary analysis function in `BAtrending`.

`compare_methods()` uses `lme4::lmer()` to fit a mixed model with differences (alt - ref) as the dependent variable. From this model, Bland-Altman statistics and treniding parameters are calculated accounting for repeated measurements. Additionally, a mixed model with averages ((alt + ref)/2) as the dependent variable is fitted. This model is used to determine between-subject and within-subject variation in the measured variable. In [Part 3](#part-3-manually-fit-the-mixed-model-and-extract-features) we demonstrate how to manually fit the mixed model and extract relevant features.

```{r}
#| results: 'hide'
#| message: false

BA_CO <- compare_methods(
  CO,
  ref_col = rv,
  alt_col = ic,
  id_col = sub
)
# Bootstrap confidence intervals.
# `nsim` should probalby be at least 1999 for real use.
# BA_CO <- add_confint(BA_CO, nsim = 999)
```

We can view the raw data (`rv` and `ic`) using `plot_BA_scatter()`. It is generally recommended to set the aspect ratio to 1 for easy comparison between the methods.

```{r}
plot_BA_scatter(BA_CO, aspect_ratio = 1)
```

The standard Bland-Altman plot is created with `plot_BA()`.

```{r}
plot_BA(BA_CO, aspect_ratio = 1)
```

`plot_BA_compine()` is a convenience function, that combines three plots, the scatter plot and Bland-Altman plots shown above and a residuals plot, that shows only the within-subject variation, which is the only variation that is relevant for trending agreement.

```{r}
plot_BA_combine(BA_CO, aspect_ratio = 1, equal_scales = TRUE)
```

Setting `equal_scales = TRUE` matches the dimensions of the residuals plot to the standard Bland-Altman plot, so we can visually compare the within-subject variation (right panel) to the total variation shown in the Bland-Altman analysis (middle panel).

On the Y-axis, we see that there is little within-subject variation in the difference between simultaneous measurements, especially compared the the total variation (Y-axis of the middle panel). By itself, this indicates good trending agreement, but we also have to consider the X-axis. The middle panel shows that a wide range of CO values are included in the study, but the right panel shows that the within-subject variation in CO is very limited. Each subject have ...

```{r}

```

# Part 2: Proportional errors

The method comparison of log-transformed measurements is performed as without the transformation. In the end, the results (e.g bias, LoA and Change LoA) can be exponentiated and interpreted on the original scale. The interpretation of these back-transformed results is that they estimate ratios between the measurements instead of differences. A relative bias of 1.10 indicates that the alternative method measures 10% more than the reference method on average. Relative limits of agreement define the interval in which ratios between future paired measurements (in a random subject) with the same alternative and reference methods are expected to fall into in 95% of cases. Relative tending precision descibe the variation around the subject-specific mean ratio between the paired measurements.

## Compare results with those from an analysis on absolute scale

To compare results from an analysis on log transformed measurements with an analysis on absolute (non-transformed) measurements, the SD of differences of $log_e$ measurements ($D_{log} = log(CO_{alt}) - log(CO_{ref})$) is comparable to the SD of differences on raw measurements ($D$) relative to the mean of the measurements:

$$
SD(D_{log}) \approx \frac{SD(D)}{mean(CO_{alt}, CO_{ref})},
$$

when $SD < \sim0.2$.

# Part 3: Manually fit the mixed model and extract features {#part-3-manually-fit-the-mixed-model-and-extract-features}