---
title: "Introduction to BAtrending"
output: 
  rmarkdown::html_vignette: default
  pdf_document:
   toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to BAtrending}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>"
)
```

# Setup

First, install the package using `pak::pak("JohannesNE/BAtrending")` or `remotes::install_github("JohannesNE/BAtrending")`.
Then, load the package.

```{r setup}
library(BAtrending)
```

# Example with absolute errors

BAtrending package includes a small sample dataset, `CO`. It contains paired measurements of cardiac output (CO) with two methods: radionuclide ventriculography (rv) and impedance cardiography (ic). The dataset was published in [Bland JM, Altman DG. (1999) Measuring agreement in method comparison studies. Statistical Methods in Medical Research 8, 135-160](https://doi.org/10.1177/096228029900800204).

The `CO` dataset has 60 measurements in 12 subjects.

```{r}
#| results: 'hide'
#| message: false

BA_CO <- compare_methods(
  CO,
  ref_col = rv,
  alt_col = ic,
  id_col = sub
)
# Bootstrap confidence intervals.
# `nsim` should probalby be at least 1999 for real use.
# BA_CO <- add_confint(BA_CO, nsim = 999)
```

`compare_methods()` is the main analysis function in `BAtrending`. Run `?compare_methods` for instructions.

`compare_methods()` uses `lme4::lmer()` to  fit a mixed model with differences (alt - ref) as the dependent variable. From this model, Bland-Altman statistics and treniding parameters are calculated accounting for repeated measurements. Additionally, a mixed model with averages ((alt + ref)/2) as the dependent variable is fitted. This model is used to determine between-subject and within-subject variation in the measured variable. In [Part 3](#part-3-manually-fit-the-mixed-model-and-extract-features) we demonstrate how to manually fit the mixed model and extract relevant features.

We can view the raw data using `plot_BA_scatter()`. It is generally recommended to set the aspect ratio to 1 for easy comparison between the methods. 

```{r}
plot_BA_scatter(BA_CO, aspect_ratio = 1)
```

The standard Bland-Altman plot is created with `plot_BA()`.

```{r}
plot_BA(BA_CO, aspect_ratio = 1)
```

 

```{r}
plot_BA_combine(BA_CO, aspect_ratio = 1)
```

```{r}

```

# Part 2: Proportional errors

The method comparison of log-transformed measurements is performed as without the transformation. In the end, the results (e.g bias, LoA and Change LoA) can be exponentiated and interpreted on the original scale. The interpretation of these back-transformed results is that they estimate ratios between the measurements instead of differences. A relative bias of 1.10 indicates that the alternative method measures 10% more than the reference method on average. Relative limits of agreement define the interval in which ratios between future paired measurements (in a random subject) with the same alternative and reference methods are expected to fall into in 95% of cases. Relative tending precision descibe the variation around the subject-specific mean ratio between the paired measurements.

## Compare results with those from an analysis on absolute scale

To compare results from an analysis on log transformed measurements with an analysis on absolute (non-transformed) measurements, the SD of differences of $log_e$ measurements ($D_{log} = log(CO_{alt}) - log(CO_{ref})$) is comparable to the SD of differences on raw measurements ($D$) relative to the mean of the measurements:

$$
SD(D_{log}) \approx \frac{SD(D)}{mean(CO_{alt}, CO_{ref})},
$$

when $SD < \sim0.2$.

# Part 3: Manually fit the mixed model and extract features