---
title: "Introduction to BAtrending"
output: 
  rmarkdown::html_vignette: default
  pdf_document:
   toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to BAtrending}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>"
)
```

# Setup

First, install the package using `pak::pak("JohannesNE/BAtrending")` or `remotes::install_github("JohannesNE/BAtrending")`.
Then, load the package.

```{r setup}
library(BAtrending)
```

# Example with absolute errors

BAtrending package includes a small sample dataset, `CO`. It contains paired measurements of cardiac output (CO) with two methods: radionuclide ventriculography (rv) and impedance cardiography (ic). The dataset was published in [Bland JM, Altman DG. (1999) Measuring agreement in method comparison studies. Statistical Methods in Medical Research 8, 135-160](https://doi.org/10.1177/096228029900800204).

The `CO` dataset has 60 measurements in 12 subjects.

```{r}
#| results: 'hide'
#| message: false

BA_CO <- compare_methods(
  CO,
  ref_col = rv,
  alt_col = ic,
  id_col = sub
)
# Bootstrap confidence intervals.
# `nsim` should probalby be at least 1999 for real use.
# BA_CO <- add_confint(BA_CO, nsim = 999)
```

`compare_methods()` is the main analysis function in `BAtrending`. Run `?compare_methods` for instructions.

`compare_methods()` uses `lme4::lmer()` to  fit a mixed model with differences (alt - ref) as the dependent variable. From this model, Bland-Altman statistics and treniding parameters are calculated accounting for repeated measurements. Additionally, a mixed model with averages ((alt + ref)/2) as the dependent variable is fitted. This model is used to determine between-subject and within-subject variation in the measured variable. In [Part 3](#part-3-manually-fit-the-mixed-model-and-extract-features) we demonstrate how to manually fit the mixed model and extract relevant features.

We can view the raw data using `plot_BA_scatter()`. It is generally recommended to set the aspect ratio to 1 for easy comparison between the methods. 

```{r}
plot_BA_scatter(BA_CO, aspect_ratio = 1)
```

The standard Bland-Altman plot is created with `plot_BA()`.

```{r}
plot_BA(BA_CO, aspect_ratio = 1)
```

 

```{r}
plot_BA_combine(BA_CO, aspect_ratio = 1)
```

```{r}

```

# Part 3: Manually fit the mixed model and extract features